A Formalization of Typed Lua

### 2015-06-13 doc/papers/paper/paper.tex
DLS paper: add preprint option for page numbers
DLS paper: reformat section on projection types, fix conclusions

\documentclass{sigplanconf}

@@

\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country}
\copyrightyear{20yy}
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm}
\doi{nnnnnnn.nnnnnnn}

\titlebanner{banner above paper title}
\preprintfooter{short description of paper}

\title{Formalization of Typed Lua}

@@

@@

\begin{figure*}[!ht]
$$
\begin{array}{rlr}
T ::= & & \textsc{first-level types:}\\
& \;\; F & \textit{first-class types}\\
& | \; \phi(F_{1},F_{2}) & \textit{filter types}\\
& | \; \pi_{i}^{x} & \textit{projection types}\\
F ::= & & \textsc{{\small first-class types:}}\\
& \;\; L & \textit{literal types}\\
& | \; B & \textit{base types}\\
& | \; \Nil & \textit{nil type}\\
& | \; \Value & \textit{top type}\\
& | \; \Any & \textit{dynamic type}\\
& | \; \Self & \textit{self type}\\
& | \; F_{1} \cup F_{2} & \textit{union types}\\
& | \; S_{1} \rightarrow S_{2} & \textit{function types}\\
& | \; \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique|open|fixed|closed} & \textit{table types}\\
& | \; x & \textit{type variables}\\
& | \; \mu x.F & \textit{recursive types}\\
%\multicolumn{3}{c}{}\\
L ::= & & \textsc{{\small literal types:}}\\
& \;\; \False \; | \; \True \; | \; {\it int} \; | \; {\it float} \; | \; {\it string} &\\
%\multicolumn{3}{c}{}\\
B ::= & & \textsc{{\small base types:}}\\
& \;\; \Boolean \; | \; \Integer \; | \; \Number \; | \; \String &\\
%\multicolumn{3}{c}{}\\
K ::= & & \textsc{{\small key types:}}\\
& \;\; L \; | \; B \; | \; \Value &\\
%\multicolumn{3}{c}{}\\
V ::= & & \textsc{{\small value types:}}\\
& \;\; F \; | \; \Const \; F &\\ 
%\multicolumn{3}{c}{}\\
S ::= & & \textsc{second-level types:}\\
& \;\; P & \textit{tuple types}\\
& | \; S_{1} \sqcup S_{2} & \textit{unions of tuple types}\\
%\multicolumn{3}{c}{}\\
P ::= & & \textsc{{\small tuple types:}}\\
%& \;\; \Void & \textit{void type}\\
& | \; F{*} & \textit{variadic types}\\
& | \; T \times P & \textit{pair types}
\end{array}
$$
\caption{Type Language}
\label{fig:typelang}
\end{figure*}

@@

@@

@@

Intuitively, a projection type $\pi_{i}^{x}$ denotes the union of
the i-th components of all the tuples in the union of tuple types
denoted by the label $x$. Unions of tuple types referenced by projections are kept in their own {\em projection environment} $\Pi$. A fresh label is generated whenever the last term
in an expression list has a union of tuple types as its type.

Notice that in the above example the local variables $q$ and $r$ are declared without 
an explicit type. The rule for typing this {\bf local} statement,
{\sc T-LocalProj}, introduces a new label in the projection
environment used for typing the {\bf if} statement:
\[
\begin{array}{c}
\mylabel{T-LocalProj}\\
\dfrac{\begin{array}{c}
	\env_{1}, \penv \vdash el:T_1 \times \ldots T_n \times P, \env_{2}, (x,S)\\
	\env_{3} = \env_{2}[id_{1} \mapsto T_1, \ldots, id_{n} \mapsto T_n]\\
	\env_{3}, \penv[x \mapsto S] \vdash s, \env_{4} \;\;\;
	n = |\;\overline{id}\;|  
	\end{array}}
{\env_{1}, \penv \vdash \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s, (\env_{4}-\{\overline{id}\})[\overline{id \mapsto \env_2(id)}]}
\end{array}
\]

The first judgement above the bar uses the auxiliary typing relation $\env_{1}, \penv \vdash el : S_{1}, \env_{2}, (x,S_{2})$.
This rule does the work of generating a fresh projection label,
and projection types associated with it, if the last term of
the expression list has a union of tuple types as its type,
as rule {\sc T-ExpListProj} below shows:
\[
\begin{array}{c}
\mylabel{T-ExpListProj}\\
\dfrac{\begin{array}{c}
	\env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
	\env_{1}, \penv \vdash me:S, \env_{n+2} \;\;\; n=|\overline{e}|\\
	S = S_1 \sqcup S_2 \;\;\;m = max(|S_1|, |S_2|)-1 \;\;\; x \notin \Pi\\
	\env_{f} = merge(\env_{1}, ..., \env_{n+2})
	\end{array}}
{\env_{1}, \penv \vdash \overline{e},me:}\\ {T_{1} \times ... \times T_{n} \times \pi_{1}^{x} \times ... \times \pi_{m}^{x} \times last(S), \env_{f}, (x,S)}
\end{array}
\]

In the above example, the combination of these two rules will bind $x$ to
$(\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*})$ in the projection environment $\Pi$,
and $q$ and $r$ respectively to $\pi_{1}^{x}$ and $\pi_{2}^{x}$
in the type environment used for typing the {\bf if} statement.

At this point, using $q$ or $r$ as an r-value will project
their types into either $\Integer \cup \Nil$ or $\Integer \cup \String$. But the more interesting case is inside the
branches of the {\bf if} statement. The particular uses the rule {\sc T-IfFilterNilProj} below, which uses flow typing:
\[
%\begin{array}{c}
%\mylabel{T-IfFilterNilProj}\\
%\dfrac{\begin{array}{c}
%       \env_{1}(id) = \pi_{i}^{x} \;\;\; \penv(x) = S\\
%       closeall(\env_{1}), \penv[x \mapsto fpt(S,\Nil,i)] \vdash s_{1}, \env_{2} \\
%       closeall(\env_{1}), \penv[x \mapsto gpt(S,\Nil,i)] \vdash s_{2}, \env_{3} \\
%       \env_{4} = closeset(\env_{1}, fav(s_{1}) \cup fav(s_{2})) \\
%       \env_{5} = openset(\env_{4}, rv(s_{1}) \cup rv(s_{2}))
%      \end{array}}
%      {\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{5}}
%\end{array}
\begin{array}{c}
\mylabel{T-IfFilterNilProj}\\
\dfrac{\begin{array}{c}
       \env_{1}(id) = \pi_{i}^{x} \;\;\; \penv(x) = S\\
       \env_{1}, \penv[x \mapsto fpt(S,\Nil,i)] \vdash s_{1}, \env_{2} \\
       \env_{1}, \penv[x \mapsto gpt(S,\Nil,i)] \vdash s_{2}, \env_{3} \\
       \env_{4} = join(\env_{2}, \env_{3}) 
      \end{array}}
      {\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{4}}
\end{array}
\]

Rule \textsc{T-IfFilterNilProj} uses the auxiliary functions \emph{fpt} and \emph{gpt}
to filter a projection $x$. Function {\em fpt} filters out $\Nil$ from the i-th
component of each tuple, removing the entire tuple if the i-th
component does not exist, or is $\Nil$ itself. Function {\em gpt}
filters the i-th component to $\Nil$ if the type includes
$\Nil$, removing the entire tuple if the i-th component does
not exist, or does not include $\Nil$.

@@

Assignment to an l-value with a projection type is forbidden. Allowing assignment
would be unsafe, as it could break the dependency among the
components of the union, as the following example shows:
\begin{verbatim}
  local q, r = idiv(a, b)
  r = "foo"
  if q then
    -- wrongly assume q and r are integers
  end
\end{verbatim}

@@

\section{Conclusion}
\label{sec:conclusion}

We have presented a formalization of the Typed Lua,
an optional type system for Lua, with a focus on rules
for two novel features of the system that type unusual
idioms present in Lua programs.

In the first idiom, records and objects are
defined through assignment to an initially empty
table. In Typed Lua's type system the type of a
table can change, either by assignment
or by using the table in a context that expects
a different (but compatible) type than the one it
has. The system tracks aliasing of table references
to be able to do this type evolution in a type-safe
way.

In the second idiom, a function that can return different
kinds of sequences of return values (for example, one
sequence for its usual path, another for error conditions)
is modeled by having a union of tuple types as its return
type. A destructuring assignment on a union of tuple types
decomposes the tuple in a way that the dependencies between
the types of each member of the union can be tracked.
Narrowing the type of one of the tuple's components with
a type predicate can then narrow the types of the other
components.

A key feature in optional type systems is usability.
Optional type systems retrofitted to an existing
language should fit the language's idioms, adding
static type safety to them. If the system is too
simple it may require too much of a change in the way
the programmers use the language. On the other hand,
if it is too complex it may overload the programmers
with types and error messages that are hard to
understand. The most challenging aspect of designing optional type systems is to find the right amount of complexity for a type system that feels natural to the programmers.

@@

We performed several
case studies to evaluate how successful we were in our goal of
providing an usable type system.
We evaluated 29 modules from 8 different case studies,
and we could give precise static types to 83\% of the 449
members that these modules export.
For half of the modules, we could give precise static types to
at least 89\% of the members from each module.

Our evaluation results show that our type system can statically
type check several Lua idioms and features, though the evaluation
also exposed several limitations of our type system.
We found that the three main limitations of our type system are
the lack of full support for overloaded functions, the lack
of parametric polymorphism, and operator overloading.
Overcoming these limitations is our major target for future work,
as it will allow us to statically type check more programs.

Unlike other optional type systems, we designed Typed Lua without
deliberate unsound parts.
However, we still do not have proofs that the novel features of
our type system are sound. We are working on a generalization
of the typing rules for evolution of table types where tagged
types can be attached to arbitrary references instead of
just local variables, and expect that a proof of soundness
will come out of this effort.

Typed Lua has a working implementation that
Lua programmers can already use as a framework to
document, test, and structure their applications,
and we already have user feedback from Lua programmers
that are using Typed Lua in their projects.
One example is ZeroBrane Studio, a Lua IDE that is
integrating Typed Lua.

Even applications where a full conversion to static
type checking is unfeasible in the current state
of the type system, or too much work, can use Typed
Lua to document the external interfaces of its
libraries, giving the benefits of static typing
checking to the users of those libraries.

>>>>

\documentclass[preprint]{sigplanconf}

@@

\conferenceinfo{DLS '15}{Month d--d, 20yy, City, ST, Country}
\copyrightyear{20yy}
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm}
\doi{nnnnnnn.nnnnnnn}

\titlebanner{Draft}
\preprintfooter{A Formalization of Typed Lua}

\title{A Formalization of Typed Lua}

@@

A key feature in optional type systems is usability.
Optional type systems retrofitted to an existing
language should fit the language's idioms, adding
static type safety to them. If the system is too
simple it may require too much of a change in the way
programmers use the language. On the other hand,
if it is too complex it may overload the programmers
with types and error messages that are hard to
understand. The most challenging aspect of designing optional type systems is to find the right amount of complexity for a type system that feels natural to the programmers.

@@

\begin{figure*}[t]
$$
\begin{array}{rlr}
T ::= & & \textsc{first-level types:}\\
& \;\; F & \textit{first-class types}\\
& | \; \phi(F_{1},F_{2}) & \textit{filter types}\\
& | \; \pi_{i}^{x} & \textit{projection types}\\
F ::= & & \textsc{{\small first-class types:}}\\
& \;\; L & \textit{literal types}\\
& | \; B & \textit{base types}\\
& | \; \mathbf{void} & \textit{void type}\\
& | \; \Nil & \textit{nil type}\\
& | \; \Value & \textit{top type}\\
& | \; \Any & \textit{dynamic type}\\
& | \; \Self & \textit{self type}\\
& | \; F_{1} \cup F_{2} & \textit{union types}\\
& | \; S_{1} \rightarrow S_{2} & \textit{function types}\\
& | \; \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique|open|fixed|closed} & \textit{table types}\\
& | \; x & \textit{type variables}\\
& | \; \mu x.F & \textit{recursive types}\\
%\multicolumn{3}{c}{}\\
L ::= & & \textsc{{\small literal types:}}\\
& \;\; \False \; | \; \True \; | \; {\it int} \; | \; {\it float} \; | \; {\it string} &\\
%\multicolumn{3}{c}{}\\
B ::= & & \textsc{{\small base types:}}\\
& \;\; \Boolean \; | \; \Integer \; | \; \Number \; | \; \String &\\
%\multicolumn{3}{c}{}\\
K ::= & & \textsc{{\small key types:}}\\
& \;\; L \; | \; B \; | \; \Value &\\
%\multicolumn{3}{c}{}\\
V ::= & & \textsc{{\small value types:}}\\
& \;\; F \; | \; \Const \; F &\\ 
%\multicolumn{3}{c}{}\\
S ::= & & \textsc{second-level types:}\\
& \;\; P & \textit{tuple types}\\
& | \; S_{1} \sqcup S_{2} & \textit{unions of tuple types}\\
%\multicolumn{3}{c}{}\\
P ::= & & \textsc{{\small tuple types:}}\\
%& \;\; \Void & \textit{void type}\\
& | \; F{*} & \textit{variadic types}\\
& | \; T \times P & \textit{pair types}
\end{array}
$$
\caption{Type Language}
\label{fig:typelang}
\end{figure*}

@@

The type {\bf void} is a type with no values, used by the
type system as a way to detect branches that are unreacheable
due to flow typing. This type is neutral with regards to
the type union operator: ${\bf void} \cup T = T \cup {\bf void} = T$ for any type $T$.

@@

\begin{figure*}[t]
\label{fig:projtypes}
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-LocalProj}\\
\dfrac{\begin{array}{c}
	\env_{1}, \penv \vdash el:T_1 \times \ldots T_n \times P, \env_{2}, (x,S)\\
	\env_{3} = \env_{2}[id_{1} \mapsto T_1, \ldots, id_{n} \mapsto T_n]\\
	\env_{3}, \penv[x \mapsto S] \vdash s, \env_{4} \;\;\;
	n = |\;\overline{id}\;|  
	\end{array}}
{\env_{1}, \penv \vdash \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s, (\env_{4}-\{\overline{id}\})[\overline{id \mapsto \env_2(id)}]}
\end{array}
\begin{array}{c}
\mylabel{T-ExpListProj}\\
\dfrac{\begin{array}{c}
	\env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
	\env_{1}, \penv \vdash me:S, \env_{n+2} \;\;\; n=|\overline{e}|\\
	S = S_1 \sqcup S_2 \;\;\;m = max(|S_1|, |S_2|)-1 \;\;\; x \notin \Pi\\
	\env_{f} = merge(\env_{1}, ..., \env_{n+2})
	\end{array}}
{\env_{1}, \penv \vdash \overline{e},me:}\\ {T_{1} \times ... \times T_{n} \times \pi_{1}^{x} \times ... \times \pi_{m}^{x} \times last(S), \env_{f}, (x,S)}
\end{array}\\\\
\begin{array}{c}
\mylabel{T-IfFilterProjThen}\\
\dfrac{\begin{array}{c}
	\env_{1}(id) = \pi_{i}^{x} \\ S_t = fonilpt(\Pi(x), i)\\ 
	finil(proj(\Pi(x), i)) = {\bf void} \\
	\env_{1}, \penv[x \mapsto S_t] \vdash s_{1}, \env_{2} 
	\end{array}}
{\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{2}}
\end{array}
\begin{array}{c}
\mylabel{T-IfFilterProjElse}\\
\dfrac{\begin{array}{c}
	\env_{1}(id) = \pi_{i}^{x} \\
	 S_e = finilpt(\Pi(x), i)\\
	fonil(proj(\Pi(x), i)) = {\bf void} \\
	\env_{1}, \penv[x \mapsto S_e] \vdash s_{2}, \env_{2} 
	\end{array}}
{\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{2}}
\end{array}
\begin{array}{c}
\mylabel{T-IfFilterProjBoth}\\
\dfrac{\begin{array}{c}
	\env_{1}(id) = \pi_{i}^{x} \\ S_t = fonilpt(\Pi(x), i) \; S_e = finilpt(\Pi(x), i)\\
	\env_{1}, \penv[x \mapsto S_t] \vdash s_{1}, \env_{2} \\
	\env_{1}, \penv[x \mapsto S_e] \vdash s_{2}, \env_{3} \\
	\env_{4} = join(\env_{2}, \env_{3}) 
	\end{array}}
{\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{4}}
\end{array}
\end{array}
\]
\caption{Rules for Projection Types}
\end{figure*}

@@

Intuitively, a projection type $\pi_{i}^{x}$ denotes the union of
the i-th components of all the tuples in the union of tuple types
denoted by the label $x$. Unions of tuple types referenced by projections are kept in their own {\em projection environment} $\Pi$. A fresh label is generated whenever the last term
in an expression list has a union of tuple types as its type.
Figure~\ref{fig:projtypes} presents the rules for introducing and filtering projection types.

Notice that in the above example the local variables $q$ and $r$ are declared without an explicit type. 
 The rule for typing this {\bf local} statement,
{\sc T-LocalProj}, introduces a new label in the projection
environment used for typing the {\bf if} statement.
The first judgement above the bar of rule {\sc T-LocalProj} uses the auxiliary typing relation $\env_{1}, \penv \vdash el : S_{1}, \env_{2}, (x,S_{2})$.

Rule {\sc T-ExpListProj} is the typing rule for this auxiliary
relation. It does the work of generating a fresh projection label,
and projection types associated with it, if the last term of
the expression list has a union of tuple types as its type.

In the above example, the combination of these two rules binds $x$ to
$(\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*})$ in the projection environment $\Pi$,
and $q$ and $r$ respectively to $\pi_{1}^{x}$ and $\pi_{2}^{x}$
in the type environment used for typing the {\bf if} statement.

At this point, using $q$ or $r$ as an r-value will project
their types into either $\Integer \cup \Nil$ or $\Integer \cup \String$. The more interesting cases are inside the
branches of the {\bf if} statement, where rule {\sc T-IfFilterNilProj} filters the whole union of tuples referenced
by the projection type.

Rule \textsc{T-IfFilterNilProj} uses the auxiliary functions \emph{fonilpt} ({\em filter-out nil in projection type}) and \emph{finilpt} ({\em filter-in nil in projection type}) to
filter the union referenced by $\pi_{i}^{x}$, respectively
filtering $\Nil$ out at every i-th component (removing
the entire tuple if it is $\Nil$) and filtering $\Nil$ in at every i-th component (removing the entire tuple if it is not $\Nil$).

@@

Rules {\sc T-IfFilterProjThen} and {\sc T-IfFilterProjElse}
respectively cover the case where the {\bf else} branch is unreachable and the case where the {\bf then} branch is unreacheable, because either the projected type of $\pi_i^x$
is not a supertype of $\Nil$ or is $\Nil$.

Assignment to an l-value with a projection type is forbidden. Allowing assignment
would be unsound, as it could break the dependency among the
components of the union, as the following example shows:
\begin{center}
	\begin{tabular}{ll}
		\multicolumn{2}{l}{$\mathbf{local} \, q, r = idiv(a, b) \, \mathbf{in}$} \\
		&  $r = ``foo" ;\; \mathbf{if} \, q \, \mathbf{then} \, s_1 \, \mathbf{else} \, s_2$
	\end{tabular}
\end{center}

The projected type of $r$ outside of the {\bf if} statement
is $\Integer \cup \String$, so the assignment seems correct,
but inside $s_1$ the projected type of $r$ would still be
$\Integer$.

@@

\section{Conclusions}
\label{sec:conclusion}

We have presented a formalization of Typed Lua,
an optional type system for Lua, with a focus on rules
for two novel features of the system that type unusual
idioms present in Lua programs.

In the first idiom, records and objects are
defined through assignment to an initially empty
table. In Typed Lua's type system the type of a
table can change, either by assignment
or by using the table in a context that expects
a different (but compatible) type from its current one.
The system tracks aliasing of table references
in order to do this type evolution in a type-safe
way.

In the second idiom, a function that can return different
kinds of tuples of return values (for example, one
tuple for its usual path, another for error conditions)
is modeled by having a union of tuple types as its return
type. A destructuring assignment on a union of tuple types
decomposes the tuple in a way that the dependencies between
the types of each member of the union can be tracked.
Narrowing the type of one of the tuple's components with
a type predicate can narrow the types of the other
components.

@@

We performed several
case studies to evaluate how successful we were in our goal of
providing an usable type system.
We evaluated 29 modules from 8 different case studies,
and we could give precise static types to 83\% of the 449
values that these modules export.
For half of the modules, we could give precise static types to
at least 89\% of the exports from each module.

Our evaluation results show that our type system can statically
type check several Lua idioms and features, though the evaluation
also exposed several limitations.
We found that the three main limitations are
the lack of full support for overloaded functions, the lack
of parametric polymorphism, and operator overloading.
Overcoming these limitations is one major target for future work,
as it will allow us to statically type check more programs.

Unlike some optional type systems, such as the type systems
of the Typescript and Dart languages, we designed Typed Lua without deliberately unsound parts.
However, we still do not have proofs that the novel features of
our type system are sound. We are working on a generalization
of the typing rules for evolution of table types where tagged
types can be attached to arbitrary references instead of
just local variables, and expect that a proof of soundness
will come out of this effort.

Typed Lua has a working implementation\footnote{Available
at \url{http://github.com/andremm/typelua}.} that
Lua programmers can already use as a framework to
document, test, and structure their applications.
Even applications where a full conversion to static
type checking is unfeasible in the current state
of the type system, or too much work, can use Typed
Lua to document the external interfaces of its
libraries, giving the benefits of static typing
checking to the users of those libraries.

We already have feedback from Lua projects
that are evaluating the use of Typed Lua, such
as the LuaRocks package manager, which is a key
piece of infrastructure for the Lua library ecosystem,
and ZeroBrane Studio, an IDE for Lua programs that is itself
written in Lua.
